Exoplanet population inference and the rate of Earth analogs \\ from noisy, incomplete catalogs

### 2014-06-03 document/ms.tex
fixing #9
pair-writing with hogg

\newcommand{\project}[1]{{\sffamily #1}}
\newcommand{\emcee}{\project{emcee}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\license}{MIT License}

@@

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}

@@

\title{%
    Exoplanet population inference \& the rate of Earth analogs
    from noisy, incomplete catalogs
}
% \title{%
%   Inferring the distribution and rate of Earth-like exoplanets \\
%   from noisy individual-planet inferences
% }

@@

Many exoplanets have been discovered and characterized; it is now possible to
study the population as a whole.
Although no true extra-Solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat larger.
Under strong assumptions, these exoplanet catalogs have been used to make an
extrapolated estimate of the probability or rate at which Sun-like stars host
Earth analogs.
Complicating these studies are that every catalog is censored by non-trivial
selection effects and detection efficiencies, and every system property
measurement (period, radius, \etc) is noisy.
Here we present a general hierarchical probabilistic framework for making
justified inferences about the occurrence rate of exoplanets, taking into
account survey completeness and, for the first time, \emph{observational
uncertainties}.
We are able to make fewer assumptions about the distribution than previous
studies; we only require that the occurrence rate be a smooth function of
period and radius (employing a Gaussian process).
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the whole population than standard procedures based
on weighting by inverse detection efficiency.
We apply the method to an existing catalog of small planet candidates around G
dwarf stars (Petigura \etal\ 2013).
We confirm a previous result that the radius distribution changes slope near
Earth radius.
We find that the rate of Earth analogs is about 0.02 (per star per natural
logarithmic bin in period and radius) with large uncertainty.
This number is much smaller than previous estimates made with the same data
but stronger assumptions.

@@

\citet{petigura} recently published constraints on this function based on
their independent analysis of the \kepler\ light curves for 42,000 Sun-like
stars.
This study was especially novel because the authors developed an independent
planet search pipeline (TERRA; \citealt{petigura-a}) and determined the
detection efficiency of their analysis empirically by injecting synthetic
signals into real light curves measured by \kepler.
The occurrence rate function determined by \citet{petigura} agreed
qualitatively with previous studies of small planets orbiting Sun-like stars
(\citealt{dong}).
In particular, both papers describe a ``flattening'' rate function (in
logarithmic radius) for planets around Earth's radius.
Even though no Earth analogs were discovered in their search, \citet{petigura}
used the small candidates that the did find to place an extrapolated
constraint on the frequency of Earth-like exoplanets, assuming a flat
occurrence rate in logarithmic period.

@@

\item the true occurrence rate function is \emph{smooth}\footnote{We give our
definition of ``smooth'' in more detail below but our model is very flexible
so this is not a strong restriction.}.

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurrence rate of exoplanets given all the light curves measured by
\kepler?}
Throughout this \paper, we use the notation $\rate_\ratepars(\entry)$ for the
occurrence rate \rate---parameterized by the parameters \ratepars---as a
function of the physical parameters \entry\ (orbital period, planetary radius,
\etc).
In this framework, the occurrence rate function can be ``parametric''---for
example, a power law---or a ``non-parametric'' function---such as a histogram
where the bin heights are the parameters \ratepars.
The light curve for a particular target $k$ is called $\data_k$ and the
\emph{true} set of physical parameters for that same object are given by
$\entry_k$.
A catalog provides a set of constraints on the parameters $\{\entry_k\}$
conditioned on the set light curves $\{\data_k\}$.

@@

In other words, a catalog is a dimensionality reduction of the raw data with
all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately 70,000 data points for the typical \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again.
This is true because the catalog provides constraints on the posterior
probability of the parameters $\{\entry_k\}$ under some choice of
\emph{interim prior} \interim
\begin{eqnarray}\eqlabel{crazypost}
p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) &=&
\frac{p(\{\data_k\}\,|\,\{\entry_k\})\,p(\{\entry_k\}\,|\,\interim)}
     {p(\{\data_k\}\,|\,\interim)} \quad.
\end{eqnarray}
The interim prior $p(\{\entry_k\}\,|\,\interim)$ is the prior
function---parameterized by \interim---that the author of the catalog choose
when making their measurements.

\dfmtodo{Can we incorporate Tim's comment: There's the additional level that
the catalog also distills detections from non-detections.  I don't know if
there's a place to state that here, I just thought of it.  Because the whole
"survey" is a collection of detections and non-detections.  Might be beyond
the scope of this paragraph; just thought I'd mention.}

@@

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate of
exoplanets that you would expect to observe taking into account the geometric
transit probability and any other detection efficiencies.
In practice, we can model the observable rate as
\begin{eqnarray}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency at \entry\ and
$\rate_\ratepars(\entry)$ is the object that we want to infer: the true
occurrence rate.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the occurrence rate as (for example) a broken power law or
a histogram.

@@

There is a very efficient algorithm called elliptical slice sampling (ESS;
\citealt{ess, ess-hyper}) for sampling the bin heights \ratepars\ from the
density in \eq{joint-posterior}.
Our implementation is adapted from Jo Bovy's BSD licensed ESS
code\footnote{\url{https://github.com/jobovy/bovy_mcmc/blob/master/bovy_mcmc/%
elliptical_slice.py}}.
To simultaneously marginalize over the hyperparameter choice, we use the
Metropolis--Hastings update from Algorithm 1 in \citet{ess-hyper}.
We tune the Metropolis--Hastings proposal by hand until we get an acceptance
fraction of $\sim0.2-0.4$ for the hyperparameters.

@@

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the true distribution of these
parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable approximation given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

For each catalog, we directly apply both the inverse-detection-efficiency
procedure as implemented by \citealt{petigura}\footnote{Our implementation
reproduces their results when applied to the published catalog.} and our
probabilistic method, marginalizing over the hyperparameters of the Gaussian
process regularization.
\Fig{smooth-results} and \fig{simulation-results} show the results of this
analysis in both cases.
In particular, the side panels compare the marginalized occurrence rate
functions in period and radius to the true functions that were used to
generate the catalogs.
\Fig{smooth-results} shows that even if the true occurrence rate is a smooth
function, the rate inferred by the inverse-detection-efficiency method can
appear to have sharp features.
In this first example---where the true distribution is well described by our
Gaussian process model---the probabilistic inference of the occurrence rate is
both more precise and accurate.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Jo Bovy (IAS),
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research builds on collaborations between astronomers and statisticians
forged during a three week workshop on ``Modern Statistical and Computational
Methods for Analysis of Kepler Data'' at SAMSI in June 2013.
This research made use of the NASA \project{Astrophysics Data System}.

@@

>>>>

\newcommand{\project}[1]{{\sffamily #1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\terra}{\project{TERRA}}
\newcommand{\license}{MIT License}

@@

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}
\newcommand{\True}{\foreign{True}}
\newcommand{\Truth}{\foreign{Truth}}

@@

\title{%
    Exoplanet population inference and the rate of Earth analogs \\
    from noisy, incomplete catalogs
}

@@

Many exoplanets have been discovered and characterized; it is now possible to
study the population as a whole.
Although no true extra-Solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat larger.
Under strong assumptions, these exoplanet catalogs have been used to make an
extrapolated estimate of the probability or rate at which Sun-like stars host
Earth analogs.
These studies are complicated by the fact that every catalog is censored by
non-trivial selection effects and detection efficiencies, and every property
(period, radius, \etc) is measured noisily.
Here we present a general hierarchical probabilistic framework for making
justified inferences about the occurrence rate of exoplanets, taking into
account survey completeness and, for the first time, \emph{observational
uncertainties}.
We are able to make fewer assumptions about the distribution than previous
studies; we only require that the occurrence rate be a smooth function of
period and radius (employing a Gaussian process).
By applying our method to synthetic catalogs, we demonstrate that it produces
more accurate estimates of the whole population than standard procedures based
on weighting by inverse detection efficiency.
We apply the method to an existing catalog of small planet candidates around G
dwarf stars (Petigura \etal\ 2013).
We confirm a previous result that the radius distribution changes slope near
Earth radius.
We find that the rate of Earth analogs is about 0.02 (per star per natural
logarithmic bin in period and radius) with large uncertainty.
This number is much smaller than previous estimates made with the same data
but stronger assumptions.

@@

\citet{petigura} recently published constraints on this function based on
their independent analysis of the \kepler\ light curves for 42,000 Sun-like
stars.
This study was especially novel because the authors developed an independent
planet search pipeline (\terra; \citealt{petigura-a}) and determined the
detection efficiency of their analysis empirically by injecting synthetic
signals into real light curves measured by \kepler.
The occurrence rate function determined by \citet{petigura} agreed
qualitatively with previous studies of small planets orbiting Sun-like stars
(\citealt{dong}).
In particular, both papers describe a ``flattening'' rate function (in
logarithmic radius) for planets around Earth's radius.
Even though no Earth analogs were discovered in their search, \citet{petigura}
used the small candidates that the did find to place an extrapolated
constraint on the frequency of Earth-like exoplanets, assuming a flat
occurrence rate in logarithmic period.

@@

\item the \True\footnote{In this \paper, we use ``\True'' to describe an
observable (for example, the exoplanet occurrence rate function) that would be
trivially measured in the limit of very high signal-to-noise data.
We use ``true'' to describe a simulation quantity with a value exactly known
to us.} occurrence rate function is \emph{smooth}\footnote{We give our
definition of ``smooth'' in more detail below but our model is very flexible
so this is not a strong restriction.}.

@@

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurrence rate of exoplanets given all the light curves measured by
\kepler?}
Throughout this \paper, we use the notation $\rate_\ratepars(\entry)$ for the
occurrence rate \rate---parameterized by the parameters \ratepars---as a
function of the physical parameters \entry\ (orbital period, planetary radius,
\etc).
In this framework, the occurrence rate function can be ``parametric''---for
example, a power law---or a ``non-parametric'' function---such as a histogram
where the bin heights are the parameters \ratepars.
The light curve for a particular target $k$ is called $\data_k$ and the
\True\ set of physical parameters for that same object are given by
$\entry_k$.
A catalog provides a set of constraints on the parameters $\{\entry_k\}$
conditioned on the set light curves $\{\data_k\}$.

@@

In other words, a catalog is a dimensionality reduction of the raw data with
all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately 70,000 data points for the typical \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again.
This is permitted because the catalog provides constraints on the posterior
probability of the parameters $\{\entry_k\}$ under some choice of
\emph{interim prior} \interim
\begin{eqnarray}\eqlabel{crazypost}
p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) &=&
\frac{p(\{\data_k\}\,|\,\{\entry_k\})\,p(\{\entry_k\}\,|\,\interim)}
     {p(\{\data_k\}\,|\,\interim)} \quad.
\end{eqnarray}
The interim prior $p(\{\entry_k\}\,|\,\interim)$ is the prior
function---parameterized by \interim---that the author of the catalog chose
when making their measurements.

@@

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate of
exoplanets that you would expect to observe taking into account the geometric
transit probability and any other detection efficiencies.
In practice, we can model the observable rate as
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency (including transit
probability) at \entry\ and $\rate_\ratepars(\entry)$ is the object that we
want to infer: the \True\ occurrence rate.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the occurrence rate as (for example) a broken power law or
a histogram.

The observed rate \obs{\rate} is a quantitative description of the rate at
which planets appear in the \citet{petigura} catalog; it is not a description
of the \True\ rate of exoplanets.
Inasmuch as the detection efficiency $\completeness(\entry)$ is calculated
correctly, the function $\rate_\ratepars(\entry)$ will represent the \True\
rate of exoplanets, at least where there is support in the data.
In practice, an estimate of the detection efficiency will not include every
decision or effect in the pipeline.
One example in this case is that the completeness function does not account
for the explicit removal of second (and third, and so on) planets.
In principle a more complete description of the detection efficiency can be
included and as this function becomes more accurate, our inferences about the
\True\ rate $\rate_\ratepars(\entry)$ will be less biased.

@@

There is a very efficient algorithm called elliptical slice sampling (ESS;
\citealt{ess, ess-hyper}) for sampling the bin heights \ratepars\ from the
density in \eq{joint-posterior}.
In practice, for problems with this specific structure, ESS outperforms more
traditional MCMC methods commonly employed in astrophysics
(e.g.,~\citealt{emcee}).
Our implementation is adapted from Jo Bovy's BSD licensed ESS
code\footnote{\url{https://github.com/jobovy/bovy_mcmc/blob/master/bovy_mcmc/%
elliptical_slice.py}}.
To simultaneously marginalize over the hyperparameter choice, we use the
Metropolis--Hastings update from Algorithm 1 in \citet{ess-hyper}.
We tune the Metropolis--Hastings proposal by hand until we get an acceptance
fraction of $\sim0.2-0.4$ for the hyperparameters.

@@

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the \True\ distribution of
these parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable approximation given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

For each catalog, we directly apply both the inverse-detection-efficiency
procedure as implemented by \citealt{petigura}\footnote{Our implementation
reproduces their results when applied to the published catalog.} and our
probabilistic method, marginalizing over the hyperparameters of the Gaussian
process regularization.
\Fig{smooth-results} and \fig{simulation-results} show the results of this
analysis in both cases.
In particular, the side panels compare the marginalized occurrence rate
functions in period and radius to the true functions that were used to
generate the catalogs.
\Fig{smooth-results} shows that even if the \True\ occurrence rate is a smooth
function, the rate inferred by the inverse-detection-efficiency method can
appear to have sharp features.
In this first example---where the true distribution is well described by our
Gaussian process model---the probabilistic inference of the occurrence rate is
both more precise and accurate.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
Ruth Angus (Oxford),
Jo Bovy (IAS), and
Eric Ford (PSU)
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research builds on collaborations between astronomers and statisticians
forged during a three week workshop on ``Modern Statistical and Computational
Methods for Analysis of Kepler Data'' at SAMSI in June 2013.
This research made use of the NASA \project{Astrophysics Data System}.

@@

\bibitem[Foreman-Mackey \etal(2012)]{emcee}
\dfmtodo{THE FUCK}
